---
title: "Storm Research"
author: "Emilie Wolf"
date: "July 11, 2017"
output: html_document
---
## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data

The data for this assignment comes in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site: [Storm Data (47Mb)](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Synopsis
According to my analysis, the data shows that tornadoes are the most harmful to human population health.

\  

## Environment
This analysis was performed using RStudio on a Macbook Pro.
```{r versions, echo=TRUE}
sessionInfo()
```
\  

## Data Processing
First we setup the environment and get the dataframe ready.
```{r setup, echo=TRUE,cache=TRUE,message=FALSE,warning=FALSE}
## Check for required packages and install if necessary
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)
if (!require("dplyr")) install.packages("dplyr")
library(dplyr)
if (!require("knitr")) install.packages("knitr")
library(knitr)

## Set global plotting theme for ggplot2
theme_set(theme_bw())
```

```{r download, echo=TRUE,cache=TRUE,message=FALSE}
## Download and import the dataset
stormURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("~/StormResearchData")) {dir.create("~/StormResearchData")}
download.file(stormURL, destfile = "~/StormResearchData/stormdata.csv.bz2", method = "curl")
storms <- read.csv("~/StormResearchData/stormdata.csv.bz2")
```

\  

## Exploratory Data Analysis

Now let's look around and play with the data.

```{r preview, echo=TRUE, results=FALSE}
## Preview the data
str(storms)
head(storms)
```

\  

#### Across the United States, which types of events are most harmful with respect to population health?

For this question, we need to group by event type and summarize injuries and fatalities.

```{r explore, echo=TRUE}
mostharmful <- storms %>% select(EVTYPE, FATALITIES, INJURIES) %>% group_by(EVTYPE) + 
        %>% summarize(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES))

# View Top 10 
head(arrange(mostharmful,desc(FATALITIES)),10)
head(arrange(mostharmful,desc(INJURIES)),10)
```
\  

As can be seen by the output, tornadoes rank #1 in both injuries and fatalities. 

\  

#### Across the United States, which types of events have the greatest economic consequences?

For this question, we need to group by event and summarize property damage and crop damage. 

```{r explore2, echo=TRUE}
## Select only the variables we are interested in
ecodmg <- select(storms, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, REMARKS)

## Remove observations with 0 damage
ecodmg <- filter(ecodmg, PROPDMG != 0 | CROPDMG != 0)

## Let's look at the multipliers
unique(ecodmg$PROPDMGEXP)
unique(ecodmg$CROPDMGEXP)
```

We know that H = 100, K = 1000, M = 1,000,000, B = 1,000,000,000, and ? = Unknown.
However, we also have a dash, a plus sign, and some numbers from 0 to 8. 
Since the documentation doesn't explain these codes, I will assume that numbers are
10^n. 

Let's dig deeper and explore the + and - signs...
```{r explore3, echo=TRUE}
ecodmg[ecodmg$PROPDMGEXP=="+" | ecodmg$PROPDMGEXP=="-",]$PROPDMG
ecodmg[ecodmg$PROPDMGEXP=="+" | ecodmg$PROPDMGEXP=="-",]$REMARKS
```
The numbers seem small, but the remarks indicate heavy damage. Perhaps the + and - signs should carry more weight. 


```{r explore4, echo=TRUE}
ecodmg[ecodmg$PROPDMGEXP == "h",]$PROPDMGEXP = 100

```

\  

## Results
